---
title: "Part 2- GSEA analysis(unused data-KEGG, Reactome, GO)"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    df_print: paged
---

## Introduction
This analysis continues the investigation of the dataset **GSE171663** from [Uryga et al. (2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE171663), focusing on the biological mechanisms driving Replicative Senescence in Vascular Smooth Muscle Cells (VSMCs).

  Building on the differential expression results from Part 1, this section focuses on **Gene Set Enrichment Analysis (GSEA)** to identify coordinated pathway changes. We will utilizing the `clusterProfiler` package to:

1.  **Query Pathway Databases:** Interrogate the **Gene Ontology (GO)**, **Reactome**, and **KEGG** databases to identify significantly enriched biological processes.
2.  **Visualize Enrichment:** Generate advanced visualizations—including GSEA enrichment plots, ridgeplots, and network heatplots—to characterize the functional landscape of cellular senescence.

*Prerequisites - DESeq2 analysis and extraction of DEG list*

## 1. Install and load the necessary libraries

```{r}
knitr::opts_chunk$set(
 warning = FALSE, message = FALSE
)
#install.packages("BiocManager")
#library(BiocManager)
#BiocManager::install("DESeq2")
#BiocManager::install("apeglm")
#BiocManager::install("tidyverse") # includes ggplot2 and dplyr
#BiocManager::install("EnhancedVolcano")
#install.packages("plotly")
#BiocManager::install("ComplexHeatmap") #alternative = pheatmap
#BiocManager::install( "AnnotationDbi" )
#install.packages("devtools")
#install.packages("Rtools")
#BiocManager::install("EnsDb.Hsapiens.v86")
# BiocManager::install( "clusterProfiler" )
# BiocManager::install( "enrichplot" )
# BiocManager::install("DOSE")
#BiocManager::install("ReactomePA", force = TRUE)
#BiocManager::install("msigdbr", force = TRUE)
#BiocManager::install("KEGGREST", force = TRUE)
# BiocManager::install("BiocUpgrade") ## you may need this
# BiocManager::install("GOSemSim", force = TRUE)
# install.packages("remotes")
# library(devtools)
# devtools::install_github("GuangchuangYu/GOSemSim")
#install.packages(c("ggraph", "ggnetwork")) #altering ggplots 
#devtools::install_github("datapplab/pathview")
#update.packages(c("lattice", "spatial"))
#install.packages("igraph")
```


```{r echo=FALSE}
#Load libraries and set input & output paths
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(RColorBrewer) # for a colourful plot
library(ComplexHeatmap)
library(clusterProfiler) # for PEA analysis
library(EnsDb.Hsapiens.v86)
library(DOSE)
library(enrichplot) # for visualisations
library(ggupset) # for visualisations
library(ReactomePA)
library(msigdbr)
library(KEGGREST)
library(httr2)
library(DBI)
library(GOSemSim)
library(lattice)
library(spatial)
library(pathview)
library(cowplot)
library(ggridges)
library(org.Hs.eg.db)
#Set directory, input and output path... It's relative to the file location
getwd()
in_path <- paste0(dirname(getwd()), "/Input/") # input path, where your input data is located
results <- paste0(dirname(getwd()), "/unused_results/") # output path, where you want your results exported to
figures <- paste0(dirname(getwd()), "/unused_figures/GSEA/") # output path, where you want your figures exported to
r_objects <- paste0(dirname(getwd()), "/unused_r_objects/") # output path, where you want your .RDS exported to
```

## 2. Import data, convert gene names to ENTREZ ID(required for clusterProfiler)

```{r}
#Read the data - used the DGE list that has matching symbols
df <- read.csv(paste0(in_path, 'DEgenes-all_symbol.csv'), row.names = 1)
#Annotate genes based on differential expression
df <- df %>% mutate(diffexpressed = case_when(
  log2FoldChange > 0 & padj < 0.05 ~ 'UP',
  log2FoldChange < 0 & padj < 0.05 ~ 'DOWN',
  padj > 0.05 ~ 'NO'
))

# Create a ranked list for GSEA analysis
#Extract the list of genes with ensembl IDs and log2FC
df_RL <- df[,2, drop = FALSE] %>% # drop = F prevents simplification into a vector
  rownames_to_column(var = "ensgene") %>%
  dplyr::rename(log2FC = log2FoldChange)

slice_head(df_RL, n=10)
#Convert Ensembl to EntrezID then extract it into a new data frame
edb <- EnsDb.Hsapiens.v86 #shortcut to call ensembldb
ENTREZ <- bitr(df_RL$ensgene, "GENEID", "ENTREZID", OrgDb = edb, drop = TRUE) #Alternative database would be org.Hs

#Join EntrezID into the results table and rank genes based on log2FC
df_RL_E <- df_RL %>%
  left_join(ENTREZ, by = c("ensgene" = "GENEID")) %>%
  dplyr::select(c(3,2)) %>%
  arrange(desc(log2FC)) %>%
  na.omit() %>%
  distinct(ENTREZID, .keep_all = TRUE) #only keep the unique IDs

sum(duplicated(df_RL_E)) #should equal 0 = no duplicates

#write.csv(df_RL_E, file = paste0(in_path,"Ranked_ENTREZ_GSEA.csv"), row.names = F) #Save the Input file for GSEA

```

## 3. GSEA GO analysis and visualization using gseaplot and ridgeplot

```{r, fig.dim = c(12,8)}
#GSEA GO analysis using the ranked list

#df_RL_E <- read.csv(paste0(in_path,"Ranked_ENTREZ.csv"), row.names = NULL) #Import list if needed

#Convert Ranked list into a vector
geneList <- df_RL_E$log2FC
names(geneList) <- as.character(df_RL_E$ENTREZID)

head(geneList, n =10)
#GO analysis
GOresults <- gseGO(
  geneList,
  ont = "BP",
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  exponent = 1,
  minGSSize = 100,
  maxGSSize = 500,
  eps = 0,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  verbose = TRUE,
  seed = FALSE,
  )

#Create results table
GO_gsea_BP <- GOresults@result

slice_head(GO_gsea_BP, n=10)
#Save the GO analysis output
#saveRDS(GOresults, file = paste0(r_objects,"GO_BP.RDS"))

#save results table
# write.csv(GO_gsea_BP, file = paste0(results,"GO_GSEA_BP.csv"), row.names = F)

#GO Plots
#pdf(file = paste0(figures, "GSEA/gseGO_BP.pdf"), width = 12, height = 8, bg = "white")
# goplot(GOresults)
```

```{r, fig.dim = c(6,4)}
#GSEA plots
#pdf(file = paste0(out_path, "GSEA/gseaGO_BP.pdf"), width = 6, height = 4, bg = "white")
gseaplot2(GOresults, geneSetID = 1, pvalue_table = F, title = GOresults$Description[1], base_size = 10)
```

```{r, fig.dim = c(8,6)}
#Ridgeplot
#pdf(file = paste0(out_path, "GSEA/ridge_gseaGO_BP.pdf"), width = 8, height = 6, bg = "white")
gr <- clusterProfiler::ridgeplot(GOresults, showCategory = 10, fill = "pvalue")
gr_modified <- gr + labs(x = "Ranked Genes", title = "GO_BP GSEA") + theme(plot.title = element_text(hjust = 0.5))
print(gr_modified)
#dev.set(which = dev.next())

#You can repeat the same thing with the MF and CC GO analysis(ont = "MF" or "CC" in gseGO)

```

## 4. KEGG GSEA analysis

```{r, fig.dim = c(12,12)}
#GSEA KEGG analysis using the ranked list
kk2 <- gseKEGG(geneList     = geneList,
               organism     = 'hsa',
               minGSSize    = 120,
               pvalueCutoff = 0.05,
               eps = 0,
               verbose      = FALSE)
head(kk2)

#Extract the results and geneSets
kk2_gsea_R <- kk2@result

#Save the GSEA KEGG analysis output
#saveRDS(kk2, file = paste0(r_objects,"KEGG_GSEA.RDS"))

#Save the results table
# write.csv(kk2_gsea_R, file = paste0(results,"KEGG_GSEA.csv"), row.names = F)

slice_head(kk2_gsea_R, n=10)
#Open KEGG page for pathway
#browseKEGG(kk2, "hsa04110") #checking the KEGG pathway online

#Creating a KEGG pathway plot with pathway package
hsa04110 <- pathview(gene.data  = geneList,
                     pathway.id = "hsa04110",
                     species    = "hsa",
                     limit      = list(gene=max(abs(geneList)), cpd=1), 
                     res = 300, 
                     key.align = "x", key.font.size = 5,
                     key.pos = "bottomright")

#Have to manually move the file to your target directory

```

## 5. Reactome GSEA analysis and visualization (GSEAplot, ridgeplot, dotplot, heatplot)
```{r}
##Reactome GSEA analysis
RC_gsea <- gsePathway(geneList, 
                 organism = "human", 
                 minGSSize    = 100,
               pvalueCutoff = 1,
               eps = 0,
               by = "DOSE",
               nPerm = 100)

#Extract results table
RC_gsea_R <- RC_gsea@result

slice_head(RC_gsea_R, n=10)
#Save Reactome GSEA output
#saveRDS(RC_gsea, file = paste0(r_objects,"Reactome_gsea.RDS"))

#save Reactome results file
# write.csv(RC_gsea@result, file = paste0(results,"Reactome_GSEA.csv"), row.names = F)

```

```{r, fig.dim = c(6,6)}
#GSEA plot
#pdf(file = paste0(figures, "GSEA/RC_GSEA.pdf"), width = 6, height = 4, bg = "white")
rc_gs <- gseaplot2(RC_gsea, geneSetID = 1, pvalue_table = F, title = RC_gsea$Description[1], base_size = 10)
print(rc_gs)

# dev.set(which = dev.next())
# dev.off()
#Ridgeplot
#pdf(file = paste0(figures, "GSEA/ridge_RC_GSEA.pdf"), width = 8, height = 6, bg = "white")
rr <- clusterProfiler::ridgeplot(RC_gsea, showCategory = 10, fill = "pvalue")
rr_modified <- rr + labs(x = "Ranked Genes", title = "Reactome GSEA") + theme(plot.title = element_text(hjust = 0.5))
print(rr_modified)
#dev.set(which = dev.next())

#Dotplot 
# pdf(file = paste0(figures, "GSEA/dot_RC_GSEA.pdf"), width = 6, height = 6, bg = "white")
rd <- clusterProfiler::dotplot(RC_gsea, x = "GeneRatio", color = "pvalue", size = "setSize", title = "Reactome GSEA", showCategory = 10)

rd_modified <- rd + ggtitle("Reactome GSEA") +
  scale_color_gradient(low = "blue", high = "red") +
  theme(plot.title = element_text(hjust = 0.5))

print(rd_modified)
#dev.set(which = dev.next())
```

```{r, fig.dim = c(30,8)}
#Heatplot

## convert gene IDs in GSEA output to Symbol
RCx<- setReadable(RC_gsea, 'org.Hs.eg.db', 'ENTREZID')

#Create a vector with logFC with names as ENTREZ from the gene List
#Function to create geneList with FC vector from unfiltered genes with FC list
geneList_vec_E <- function(x){
  logFC <- x$log2FC
  names(logFC) <- x$ENTREZID
  return(logFC)
}

#Convert genelist into symbol list from entrez and extract into a vector(all unfiltered genes)
#Check keytypes in org.Hs.eg.db- keytypes(org.Hs.eg.db)
geneList_vec_S <- function(x){
  EID <-as.character(x$ENTREZID) #Character vector with ENTREZID for mapping
  x$SYMBOL <-mapIds(org.Hs.eg.db, keys = EID, keytype = "ENTREZID", column = "SYMBOL", multiVals = "first") 
  logFC <- x$log2FC
  names(logFC) <- x$SYMBOL
  logFC <- logFC[!is.na(names(logFC))]
  return(logFC)
}

geneList_E <- geneList_vec_E(df_RL_E)
geneList_S <- geneList_vec_S(df_RL_E)

head(geneList_S, n=10)

#pdf(file = paste0(figures, "GSEA/heat_RC_GSEA.pdf"), width = 35, height = 10, bg = "white")
p8 <- heatplot(RCx, foldChange=geneList_S, symbol = "dot", showCategory=5) + viridis::scale_fill_viridis(name = "log2FC")

p8_modified <- p8 + ggtitle("Reactome GSEA") +
  theme(plot.title = element_text(hjust = 0.5))
print(p8_modified)
#dev.set(which = dev.next())

```

```{r sessionInfo}
sessionInfo() 

```