---
title: "Part 2-Gene Set Enrichment Analysis(WP)"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    df_print: paged
---

## Introduction
This analysis continues the investigation of the dataset **GSE171663** from [Uryga et al. (2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE171663), focusing on the biological mechanisms driving Replicative Senescence in Vascular Smooth Muscle Cells (VSMCs).

Building on the differential expression results from Part 1, this section focuses on **Gene Set Enrichment Analysis (GSEA)** to identify coordinated pathway changes. We will utilize the `clusterProfiler` package to:

1.  **Query the WikiPathways Database:** Perform GSEA specifically against the **WikiPathways** database to identify significantly enriched biological processes and signaling cascades.
2.  **Visualize Enrichment:** Generate advanced visualizations—including GSEA enrichment plots, ridgeplots, and heatplots—to characterize the functional landscape of the senescent transcriptome.

 *Prerequisites - DESeq2 analysis and extraction of DEG list*

## 1. Install and load the necessary libraries
```{r}
knitr::opts_chunk$set(
 warning = FALSE, message = FALSE
)
#install.packages("BiocManager")
#library(BiocManager)
#BiocManager::install("DESeq2")
#BiocManager::install("apeglm")
#BiocManager::install("tidyverse") # includes ggplot2 and dplyr
#BiocManager::install("EnhancedVolcano")
#install.packages("plotly")
#BiocManager::install("ComplexHeatmap") #alternative = pheatmap
#BiocManager::install( "AnnotationDbi" )
#install.packages("devtools")
#install.packages("Rtools")
#BiocManager::install("EnsDb.Hsapiens.v86")
# BiocManager::install( "clusterProfiler" )
# BiocManager::install( "enrichplot" )
# BiocManager::install("DOSE")
#BiocManager::install("ReactomePA", force = TRUE)
#BiocManager::install("msigdbr", force = TRUE)
#BiocManager::install("KEGGREST", force = TRUE)
# BiocManager::install("BiocUpgrade") ## you may need this
# BiocManager::install("GOSemSim", force = TRUE)
# install.packages("remotes")
# library(devtools)
# devtools::install_github("GuangchuangYu/GOSemSim")
#install.packages(c("ggraph", "ggnetwork")) #altering ggplots 
#devtools::install_github("datapplab/pathview")
#update.packages(c("lattice", "spatial"))
#install.packages("igraph")
```


```{r echo=FALSE}
#Load libraries and set input & output paths
library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(RColorBrewer) # for a colourful plot
library(ComplexHeatmap)
library(clusterProfiler) # for PEA analysis
library(EnsDb.Hsapiens.v86)
library(DOSE)
library(enrichplot) # for visualisations
library(ggupset) # for visualisations
library(ReactomePA)
library(msigdbr)
library(KEGGREST)
library(httr2)
library(DBI)
library(GOSemSim)
library(lattice)
library(spatial)
library(pathview)
library(cowplot)
library(ggridges)
library(org.Hs.eg.db)

#Set directory, input and output path
getwd()

#Create paths for your directories
in_path <- paste0(dirname(getwd()), "/Input/") #For input files
results <- paste0(getwd(), "/Results/") #For results and tables
figures <- paste0(dirname(getwd()), "/Figures/") #For main figures
r_objects <- paste0(getwd(), "/r_objects/") # output path, where you want your .RDS exported to
```


## 2. Import data, convert gene names to ENTREZ ID(required for clusterProfiler)
```{r}
#Read the data - used the DGE list that has matching symbols
df <- read.csv(paste0(in_path, 'DEgenes-all_symbol.csv'), row.names = 1)
#Annotate genes based on differential expression
df <- df %>% mutate(diffexpressed = case_when(
  log2FoldChange > 0 & padj < 0.05 ~ 'UP',
  log2FoldChange < 0 & padj < 0.05 ~ 'DOWN',
  padj > 0.05 ~ 'NO'
))

# Create a ranked list for GSEA analysis
#Extract the list of genes with ensembl IDs and log2FC
df_RL <- df[,2, drop = FALSE] %>% # drop = F prevents simplification into a vector
  rownames_to_column(var = "ensgene") %>%
  dplyr::rename(log2FC = log2FoldChange)

slice_head(df_RL, n=10)
#Convert Ensembl to EntrezID then extract it into a new data frame
edb <- EnsDb.Hsapiens.v86 #shortcut to call ensembldb
ENTREZ <- bitr(df_RL$ensgene, "GENEID", "ENTREZID", OrgDb = edb, drop = TRUE) #Alternative database would be org.Hs

#Join EntrezID into the results table and rank genes based on log2FC
df_RL_E <- df_RL %>%
  left_join(ENTREZ, by = c("ensgene" = "GENEID")) %>%
  dplyr::select(c(3,2)) %>%
  arrange(desc(log2FC)) %>%
  na.omit() %>%
  distinct(ENTREZID, .keep_all = TRUE) #only keep the unique IDs

sum(duplicated(df_RL_E)) #should equal 0 = no duplicates

#write.csv(df_RL_E, file = paste0(results,"Ranked_ENTREZ_GSEA.csv"), row.names = F) #Input file for GSEA

```

## 3. Create a gene list for GSEA analysis

```{r, fig.dim = c(12,8)}
#GSEA GO analysis using the ranked list

#df_RL_E <- read.csv(paste0(results,"Ranked_ENTREZ.csv"), row.names = NULL) #Import list if needed

#Convert Ranked list into a vector
geneList <- df_RL_E$log2FC
names(geneList) <- as.character(df_RL_E$ENTREZID)

head(geneList, n =10)
```

## 4. WikiPathways GSEA analysis and visualization (GSEAplot, ridgeplot, dotplot, heatplot) -->
```{r, fig.dim = c(6,4)}
###Wikipathways GSEA analysis 
WP_gsea <- gseWP(geneList, 
                 organism = "Homo sapiens", 
                 minGSSize    = 100,
               pvalueCutoff = 1,
               eps = 0,
               by = "DOSE",
               nPerm = 100)


WP_gsea_R <- WP_gsea@result

slice_head(WP_gsea_R, n=10)
#Save all GSEA output
#saveRDS(WP_gsea, file = paste0(r_objects,"Wikipathways/WP_gsea.RDS"))

#save the results file
#write.csv(WP_gsea_R, file = paste0(results,"Wikipathways/Wikipathways_GSEA.csv"), row.names = F)

#GSEA plot
# pdf(file = paste0(figures, "WP_GSEA.pdf"), width = 6, height = 4, bg = "white")
wp_gs <- gseaplot2(WP_gsea, geneSetID = 1, pvalue_table = F, title = WP_gsea$Description[1], base_size = 10)
print(wp_gs)

#dev.set(which = dev.next())
```

```{r, fig.dim = c(6,6)}
#Ridgeplot
# pdf(file = paste0(figures, "ridge_WP_GSEA.pdf"), width = 8, height = 6, bg = "white")
wr <- clusterProfiler::ridgeplot(WP_gsea, showCategory = 10, fill = "pvalue")
wr_modified <- wr + labs(x = "Ranked Genes", , title = "WikiPathways GSEA") + theme(plot.title = element_text(hjust = 0.5))
print(wr_modified)
dev.set(which = dev.next())


# #Save as png
# ggsave(
#   filename = paste0(figures,"WP_ridge.png"),
#   plot = wr_modified,
#   width = 8,
#   height = 6,
#   dpi = 300, # <--- This is the key. 300 is "Print Quality"
#   bg = "white" 
# )

#Dotplot 
# pdf(file = paste0(figures, "dot_WP_GSEA.pdf"), width = 6, height = 6, bg = "white")
wd <- clusterProfiler::dotplot(WP_gsea, x = "GeneRatio", color = "pvalue", size = "setSize", title = "WikiPathways GSEA", showCategory = 10)

wd_modified <- wd + ggtitle("WikiPathways GSEA") +
  scale_color_gradient(low = "blue", high = "red") +
  theme(plot.title = element_text(hjust = 0.5))

print(wd_modified)
#dev.set(which = dev.next())

```

```{r, fig.dim = c(30,8)}
#Heatplot

## convert gene IDs in GSEA output to Symbol
WPx<- setReadable(WP_gsea, 'org.Hs.eg.db', 'ENTREZID')


#Create a vector with logFC with names as ENTREZ from the gene List
#Function to create geneList with FC vector from unfiltered genes with FC list
geneList_vec_E <- function(x){
  logFC <- x$log2FC
  names(logFC) <- x$ENTREZID
  return(logFC)
}

#Convert genelist into symbol list from entrez and extract into a vector(all unfiltered genes)
#Check keytypes in org.Hs.eg.db- keytypes(org.Hs.eg.db)
geneList_vec_S <- function(x){
  EID <-as.character(x$ENTREZID) #Character vector with ENTREZID for mapping
  x$SYMBOL <-mapIds(org.Hs.eg.db, keys = EID, keytype = "ENTREZID", column = "SYMBOL", multiVals = "first") 
  logFC <- x$log2FC
  names(logFC) <- x$SYMBOL
  logFC <- logFC[!is.na(names(logFC))]
  return(logFC)
}

geneList_E <- geneList_vec_E(df_RL_E)
geneList_S <- geneList_vec_S(df_RL_E)

head(geneList_S, n=10)
#pdf(file = paste0(figures, "heat_WP_GSEA.pdf"), width = 35, height = 10, bg = "white")
p8 <- heatplot(WPx, foldChange=geneList_S, symbol = "dot", showCategory=5) + viridis::scale_fill_viridis(name = "log2FC")

p8_modified <- p8 + ggtitle("WikiPathways GSEA") +
  theme(plot.title = element_text(hjust = 0.5))
print(p8_modified)
#dev.set(which = dev.next())

# Save as png
ggsave(
  filename = paste0(figures,"WP_heatplot.png"),
  plot = p8_modified,
  width = 35,
  height = 10,
  dpi = 300,   # <--- This is the key. 300 is "Print Quality"
  bg = "white"
)

#Heatplot with subsets of genes(up or down)
# up <- sort(geneList_S, decreasing = TRUE)[1:30]
# geneList_UP <- geneList_S[up]
# p9 <- heatplot(WPx, foldChange=geneList_UP, symbol = "dot", showCategory=4) + viridis::scale_fill_viridis(name = "log2FC")

# p9_modified <- p9 + ggtitle("WikiPathways GSEA") +
#   theme(plot.title = element_text(hjust = 0.5))
# print(p9_modified)

#down genes
# dn <- sort(geneList_S, decreasing = FALSE)[1:30]
# geneList_dn <- geneList_S[dn]

```

```{r sessionInfo}
sessionInfo() 

```


